---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# spacefinder

<!-- badges: start -->
<!-- badges: end -->

## Overview

`spacefinder` identifies promising subspaces within hyperparameter spaces by fitting geometric regions (hyperrectangles and ellipsoids) to high-performing configurations from hyperparameter tuning benchmarks.

**Key features:**

- Three learner types: Box (axis-aligned), Polygon (oriented), and Ellipsoid
- Handles categorical hyperparameters by fitting separate subspaces per level
- Beta density estimation for probabilistic sampling (`augment()`)
- Visualization tools for fitted subspaces (`autoplot()`)
- Works with multi-task benchmark data

## Installation

Install the development version from GitHub:
```r
# install.packages("pak")
pak::pak("NikoGerman/spacefinder")
```

## Quick Start
```{r example, message=FALSE}
library(spacefinder)

# Load example data
data(benchmark_data)

# Create task
task <- TaskSubspace$new(
  data = benchmark_data,
  target_measure = "auc",
  hps = c("learning_rate", "max_depth")
)

# Fit axis-aligned hyperrectangle
learner <- LearnerSubspaceBox$new(task)
learner$train(q_val = 0.9)

# View fitted bounds
coef(learner)
```
```{r plot, fig.height=5, fig.width=7}
# Visualize fitted subspace
ggplot2::autoplot(learner)
```

## Documentation

Learn more about spacefinder:

- `vignette("getting-started", package = "spacefinder")` - Basic workflow and concepts
- `vignette("learner-comparison", package = "spacefinder")` - Comparing Box, Polygon, and Ellipsoid learners  
- `vignette("categorical-hyperparameters", package = "spacefinder")` - Working with categorical variables
- `vignette("density-estimation", package = "spacefinder")` - Probabilistic modeling with `augment()`

Or browse online at <https://nikogerman.github.io/spacefinder/>

## Citation

If you use spacefinder in your research, please cite:
```
@Manual{,
  title = {spacefinder: Subspace Learning for Hyperparameter Optimization},
  author = {Nikolai German},
  year = {2025},
  note = {R package version 0.2.0},
  url = {https://github.com/NikoGerman/spacefinder},
}
```

## License

Apache License v2.0
