---
title: "Working with Categorical Hyperparameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Categorical Hyperparameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 3,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(spacefinder)
library(data.table)
library(ggplot2)
data(benchmark_data)
```

## Introduction

Many hyperparameter optimization scenarios involve categorical hyperparameters like optimizer type, activation function, or kernel choice. The `spacefinder` package handles these by fitting separate subspaces for each categorical level.

## The Data

The `benchmark_data` includes three optimizers: SGD, Adam, and RMSprop. Let's explore how they differ:
```{r explore_optimizers}
# Performance by optimizer
benchmark_data[, .(
  mean_auc = mean(auc),
  median_auc = median(auc),
  sd_auc = sd(auc),
  max_auc = max(auc),
  n = .N
), by = optimizer]

# Visualize distributions
ggplot(benchmark_data, aes(x = auc, fill = optimizer)) +
  geom_density(alpha = 0.6) +
  theme_minimal() +
  labs(title = "AUC Distribution by Optimizer",
       x = "AUC", y = "Density") +
  theme(legend.position = "top")
```

Notice that Adam tends to achieve slightly higher performance than SGD and RMSprop.

## Creating a Task with Categorical Hyperparameters

Include the categorical hyperparameter in the task definition:
```{r create_task}
# Explicit specification
task <- TaskSubspace$new(
  data = benchmark_data,
  target_measure = "auc",
  hps = c("learning_rate", "max_depth"),
  cat_hps = "optimizer"
)

# Formula interface and convenience function
task <- as_task_subspace(
  data = benchmark_data,
  formula = auc ~ (learning_rate + max_depth) * optimizer
)
```

The `* optimizer` in the formula indicates that separate subspaces should be fitted for each optimizer type.

## Fitting Separate Subspaces

Train a learner - it will automatically fit one subspace per optimizer:
```{r train_learner}
learner <- LearnerSubspaceBox$new(task)
learner$train(q_val = 0.9)
```

## Examining Results per Category

### Coefficients

The coefficients now show separate bounds for each optimizer:
```{r coef}
bounds <- coef(learner)
print(bounds)
```

Notice how different optimizers have different optimal ranges!
```{r compare_bounds}
# Compare learning rate ranges
bounds[hyperparameter == "learning_rate"] |>
  ggplot(aes(optimizer, color = optimizer)) +
  geom_errorbar(aes(ymin = min, ymax = max), width = 0.2, linewidth = 1) +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Optimal Learning Rate Range by Optimizer",
       y = "Learning Rate (log scale)", x = "Optimizer") +
  theme(legend.position = "none")
```

### Summary

The summary shows fitting statistics for each optimizer:
```{r summary}
summary(learner)
```

Each optimizer gets its own row in the status table, showing:
- Number of observations used
- Performance across tasks

### Visualization

The autoplot function creates separate visualizations for each optimizer:
```{r visualize}
lapply(autoplot(learner), \(plot) plot + scale_x_continuous(limits = c(0, .01)))
```

Each panel shows:
- The fitted subspace for that optimizer
- Top configurations for that optimizer (orange crosses)
- All configurations for that optimizer (gray points)

## Comparing Optimizers

Let's analyze the differences between optimizers:
```{r compare_subspaces}
# Extract top configs per optimizer
top_configs_summary <- learner$top_configs[, .(
  mean_lr = mean(learning_rate),
  median_lr = median(learning_rate),
  mean_depth = mean(max_depth),
  median_depth = median(max_depth),
  mean_auc = mean(auc),
  n = .N
), by = optimizer]

print(top_configs_summary)
```
```{r visualize_comparison, fig.width=8, fig.height=5}
# Scatter plot colored by optimizer
ggplot(learner$top_configs, aes(x = learning_rate, y = max_depth, 
                                 color = optimizer, shape = optimizer)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Top Configurations by Optimizer",
       x = "Learning Rate (log scale)", y = "Max Depth") +
  theme(legend.position = "top")
```

## Next Steps

- **`vignette("learner-comparison")`**: Compare Box, Polygon, and Ellipsoid learners
- **`vignette("density-estimation")`**: Add probabilistic density with `augment()`
```{r session_info}
sessionInfo()
```
