---
title: "Beta Density Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beta Density Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(spacefinder)
library(data.table)
library(ggplot2)
library(generics)
data("benchmark_data", package = "spacefinder")
```

## Introduction

After fitting a subspace, you can add probabilistic density information using `augment()`. This fits Beta distributions to each hyperparameter dimension, allowing you to:

- Sample new configurations weighted toward better regions
- Estimate probability densities within the subspace
- Understand which parts are most promising

**Note:** `augment()` is only available for **Box** and **Polygon** learners, not Ellipsoid.

## Basic Usage
```{r fit_density}
# Train a box learner
task <- TaskSubspace$new(
  data = benchmark_data,
  target_measure = "auc",
  hps = c("learning_rate", "max_depth")
)

learner <- LearnerSubspaceBox$new(task)
learner$train(q_val = 0.9)

# Fit beta densities
densities <- augment(learner)
print(densities)
```

The `augment()` function returns fitted Beta distribution parameters (α, β) for each hyperparameter.

## Understanding Beta Parameters
```{r interpret_beta}
# Extract parameters for learning_rate
lr_params <- densities[parameter == "learning_rate"]
alpha <- lr_params$alpha
beta <- lr_params$beta

cat("Learning rate Beta(", alpha, ", ", beta, ")\n", sep = "")
cat("Mode at:", (alpha - 1) / (alpha + beta - 2), "\n")
```

- **α = β = 1**: Uniform distribution (all values equally likely)
- **α > 1, β > 1**: Peak in the interior (preferred region)
- **α > β**: Peak toward 1 (prefers higher values in [0,1])
- **α < β**: Peak toward 0 (prefers lower values in [0,1])

## Visualizing Densities
```{r visualize_density, fig.width=7, fig.height=4}
# Plot beta densities
x <- seq(0, 1, length.out = 200)

plot_data <- rbindlist(lapply(densities$parameter, function(hp) {
  params <- densities[parameter == hp]
  data.table(
    parameter = hp,
    x = x,
    density = dbeta(x, params$alpha, params$beta)
  )
}))

ggplot(plot_data, aes(x = x, y = density)) +
  geom_line(color = "#2E86AB", linewidth = 1) +
  facet_wrap(~parameter, scales = "free_y") +
  theme_minimal() +
  labs(title = "Fitted Beta Densities",
       subtitle = "Higher density = more promising region",
       x = "Normalized position in subspace [0,1]",
       y = "Density")
```

Higher density indicates regions where good configurations are more concentrated.

## Sampling New Configurations

You can sample from the fitted distributions to generate new hyperparameter configurations:
```{r sampling}
set.seed(123)
n_samples <- 5

# Get bounds
bounds <- coef(learner)
lr_bounds <- bounds[hyperparameter == "learning_rate"]
depth_bounds <- bounds[hyperparameter == "max_depth"]

# Sample from beta distributions (in [0,1])
lr_params <- densities[parameter == "learning_rate"]
depth_params <- densities[parameter == "max_depth"]

samples_unit <- data.table(
  learning_rate_unit = rbeta(n_samples, lr_params$alpha, lr_params$beta),
  max_depth_unit = rbeta(n_samples, depth_params$alpha, depth_params$beta)
)

# Transform to original scale
samples <- data.table(
  learning_rate = samples_unit$learning_rate_unit * 
    (lr_bounds$max - lr_bounds$min) + lr_bounds$min,
  max_depth = samples_unit$max_depth_unit * 
    (depth_bounds$max - depth_bounds$min) + depth_bounds$min
)

print(samples)
```

These sampled configurations are weighted toward the most promising regions.

## Regularization

By default, `augment()` uses `regularize = TRUE` to avoid U-shaped densities:
```{r regularization}
# With regularization (default)
densities_reg <- augment(learner, regularize = TRUE)

# Without regularization
densities_unreg <- augment(learner, regularize = FALSE)

cat("With regularization:\n")
print(densities_reg[, .(parameter, alpha, beta)])

cat("\nWithout regularization:\n")
print(densities_unreg[, .(parameter, alpha, beta)])
```

Regularization enforces α ≥ 1 and β ≥ 1, ensuring the mode exists in the interior of [0,1].

## Categorical Hyperparameters

When using categorical hyperparameters, separate densities are fitted per level:
```{r categorical}
task_cat <- TaskSubspace$new(
  data = benchmark_data,
  target_measure = "auc",
  hps = c("learning_rate", "max_depth"),
  cat_hps = "optimizer"
)

learner_cat <- LearnerSubspaceBox$new(task_cat)
learner_cat$train(q_val = 0.9)

densities_cat <- augment(learner_cat)
print(densities_cat)
```

Each optimizer gets its own Beta distribution parameters for each hyperparameter.

## Polygon Learner

`augment()` also works with Polygon learners:
```{r polygon}
learner_polygon <- LearnerSubspacePolygon$new(task)
learner_polygon$train(q_val = 0.9, lambda = 0.1)

densities_polygon <- augment(learner_polygon)
print(densities_polygon)
```

The process is the same: data is transformed to the unit cube, then Beta distributions are fitted.

## Summary

- `augment()` adds Beta density parameters to Box and Polygon learners
- Higher density indicates more promising regions
- Use fitted densities to sample new configurations
- Regularization (default) prevents U-shaped densities
- Works with categorical hyperparameters

## Next Steps

- `vignette("learner-comparison")`: Compare Box, Polygon, and Ellipsoid learners
- `vignette("categorical-hyperparameters")`: Handle categorical variables
```{r session_info}
sessionInfo()
```
