---
title: "Getting Started with spacefinder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spacefinder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```
```{r setup}
library(spacefinder)
library(data.table)
library(ggplot2)
set.seed(42)
```

## Introduction

The `spacefinder` package helps you identify promising hyperparameter subspaces from benchmark data. Instead of exploring the entire hyperparameter space, you can focus on regions that consistently produce high-performing models.

## Example Data

## Example Data

We'll use the included `benchmark_data` dataset, which contains synthetic hyperparameter tuning results:
```{r load_data}
data("benchmark_data", package = "spacefinder")
head(benchmark_data)

# Summary statistics
summary(benchmark_data$auc)
cat("\nConfigurations with AUC > 0.9:", sum(benchmark_data$auc > 0.9), "\n")
```

The dataset includes:
- 5 tasks (different datasets)
- 200 configurations per task
- Hyperparameters: `learning_rate`, `max_depth`, `optimizer`
- Performance: `auc` (Area Under ROC Curve)

## Creating a Task

Define a subspace task specifying the data, target measure, and hyperparameters:
```{r create_task}
task <- TaskSubspace$new(
  data = benchmark_data,
  target_measure = "auc",
  hps = c("learning_rate", "max_depth")
)

# Alternative: formula interface
task <- TaskSubspace$new(
  data = benchmark_data,
  formula = auc ~ (learning_rate + max_depth)
)
```

## Training a Learner

We'll use the Box learner, which fits axis-aligned hyperrectangles (simple bounds per hyperparameter):
```{r train_learner}
# Initialize learner
learner <- LearnerSubspaceBox$new(task)

# Train on top 10% of configurations
learner$train(q_val = 0.9)
```

The `q_val = 0.9` parameter means we use only configurations with performance above the 90th percentile (top 10%) for each task.

## Extracting Results

### Coefficients

Get the fitted hyperparameter bounds:
```{r coef}
bounds <- coef(learner)
print(bounds)
```

The learned optimal ranges:
- `learning_rate`: between `r sprintf("%.4f", bounds$min[1])` and `r sprintf("%.4f", bounds$max[1])`
- `max_depth`: between `r bounds$min[2]` and `r bounds$max[2]`

### Summary

Get a comprehensive overview:
```{r summary}
summary(learner)
```

The summary shows:
- **Summary**: Task information
- **Coefficients**: Fitted bounds
- **Status**: Number of observations used for fitting

### Visualization

Visualize the fitted subspace:
```{r visualize, fig.width=8, fig.height=6}
autoplot(learner)
```

The plot shows:
- **Blue rectangle**: Fitted subspace bounds
- **Orange crosses**: Top-performing configurations used for fitting
- **Gray points**: All configurations (background)

## Next Steps

Learn more about spacefinder:

- **`vignette("categorical-hyperparameters")`**: Handle categorical variables (e.g., optimizers)
- **`vignette("learner-comparison")`**: Compare Box, Polygon, and Ellipsoid learners
- **`vignette("density-estimation")`**: Add probabilistic density with `augment()`

## Session Info
```{r session_info}
sessionInfo()
```
