---
title: "Comparing Subspace Learners"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Subspace Learners}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(spacefinder)
library(data.table)
library(ggplot2)
data("benchmark_data", package = "spacefinder")
```

## Introduction

The `spacefinder` package offers three learner types that differ in how they represent promising hyperparameter regions:

- **Box**: Axis-aligned hyperrectangles (simple bounds per hyperparameter)
- **Polygon**: Oriented hyperrectangles (can rotate to capture correlations)
- **Ellipsoid**: Ellipsoids (smooth, curved boundaries)

Let's see how they compare on the same data.

## Training All Three Learners
```{r create_task}
# Create task
task <- TaskSubspace$new(
  data = benchmark_data,
  target_measure = "auc",
  hps = c("learning_rate", "max_depth")
)

# Train all three learners with identical settings
learner_box <- LearnerSubspaceBox$new(task)
learner_box$train(q_val = 0.9, lambda = 0.1)

learner_polygon <- LearnerSubspacePolygon$new(task)
learner_polygon$train(q_val = 0.9, lambda = 0.1)

learner_ellipsoid <- LearnerSubspaceEllipsoid$new(task)
learner_ellipsoid$train(q_val = 0.9, lambda = 0.1)
```

## Visual Comparison

The best way to understand the differences is to visualize them:
```{r visual_comparison, fig.width=6, fig.height=4.8}
p_box <- autoplot(learner_box) + 
  ggtitle("Box: Axis-aligned") +
  scale_x_continuous(limits = c(0, 0.01), breaks = seq(0, 0.01, by = 0.0025), guide = guide_axis(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

p_polygon <- autoplot(learner_polygon) + 
  ggtitle("Polygon: Oriented") +
  scale_x_continuous(limits = c(0, 0.01), breaks = seq(0, 0.01, by = 0.0025), guide = guide_axis(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

p_ellipsoid <- autoplot(learner_ellipsoid) + 
  ggtitle("Ellipsoid: Smooth") +
  scale_x_continuous(limits = c(0, 0.01), breaks = seq(0, 0.01, by = 0.0025), guide = guide_axis(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 10))

# Combine plots
patchwork::wrap_plots(p_box, p_polygon, p_ellipsoid, ncol = 3)
```


**Key differences:**

- **Box**: Simple rectangle aligned with axes - easiest to interpret
- **Polygon**: Can rotate to better fit the data - captures diagonal patterns
- **Ellipsoid**: Smooth curved boundary - most flexible fit

## Coefficients

### Box: Simple Bounds
```{r box_coef}
coef(learner_box)
```

Box learners give you direct bounds: "use learning_rate between X and Y".

### Polygon and Ellipsoid: Transformation Matrices
```{r other_coefs}
# Polygon
coef(learner_polygon)

# Ellipsoid  
coef(learner_ellipsoid)
```

Polygon and Ellipsoid use transformation matrices **A** and **b**. These are harder to interpret but allow more flexible shapes.

## Outliers
```{r outliers}
cat("Outliers identified:\n")
cat("  Box:", nrow(outliers(learner_box)), "\n")
cat("  Polygon:", nrow(suppressMessages(outliers(learner_polygon))), "\n")
cat("  Ellipsoid:", nrow(suppressMessages(outliers(learner_ellipsoid))), "\n")
```

Different learners may identify different configurations as outliers based on their geometry.

## Next Steps

- `vignette("density-estimation")`: Add probabilistic density with `augment()`
- `vignette("categorical-hyperparameters")`: Handle categorical variables
```{r session_info}
sessionInfo()
```
